A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-23 main.tex
4.1

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. \anna{Qualitatively, this is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by more discrete, nonoverlapping genetic subpopulations, which more directly correspond to the distinct fixed environmental effects than do the continuous and blended subpopulations of the 1D Linear Admixture data. In other words, a greater proportion of the environmental effect is in the column space of the projection matrix of $\bX$ for the Empirical and Independent Subpopulations data than for the 1D Linear Admixture data. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the ratio \eqref{eqn:ratio} introduced in Section \ref{sec:sim_env_conf}. $||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2$, is equal to 0.98615, 0.64325, and 0.46412 for the Independent Subpopulations, Empirical, and 1D Linear Admixture data, respectively. As such there is a greater amount of confounding and estimation bias which the PC-Lasso and LMM-Lasso methods are able to correct for in the Independent Subpopulations, while there is less confounding which can be corrected for in the 1D Linear Admixture data.}

\anna{\textbf{check this value for sampsizes of 588 for admixture and independent...maybe even diff subpop sizes? at least to pin down the issue - more comparable with same sampesize and subpop sizes hmm}}

@@

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean TPR and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data. \anna{Again, this is due to the smaller proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ for the 1D Linear Admixture data, which results in less confounded SNP-phenotype associations to begin with.}

>>>>

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data.

Qualitatively, this is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by more discrete, nonoverlapping genetic subpopulations, which more directly correspond to the nonoverlapping environmental effects than do the continuous and blended subpopulations of the 1D Linear Admixture data. In other words, a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$ for the Empirical and Independent Subpopulations data than for the 1D Linear Admixture data. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. CP $= 0.46, 0.99, 0.64$ for the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-Lasso and LMM-Lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-Lasso and LMM-Lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

\anna{\textbf{check this value for sampsizes of 588 for admixture and independent...maybe even diff subpop sizes? at least to pin down the issue - more comparable with same sampesize and subpop sizes but still different...}}

\anna{with the same sample size and subpopulation structure as the vanderwas data, I can replicate the CP for the indepenent subpopulations data when simulating using a lower Fst value (as defined in the ochoa and storey papers: the mean inbreeding coefficient of the individuals’ local populations relative to their last common ancestral population. This FST naturally yields a useful pairwise FST between individuals.) So, this suggests that the greater improvement in the semi-empirical data is because there is less cryptic relatedness/inbreeding (?????) still thinking about this...}

@@

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean TPR and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the smaller proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ for the 1D Linear Admixture data, which results in less confounded SNP-phenotype associations to begin with.
